---
title: "Data Cleaning and Integration in Official Statistics"
subtitle: "Part 1 Project - Ilaria Bombelli"
author: "Claire du Campe de Rosamel"
format:
  beamer:
    theme: Madrid
    fontsize: 9pt
---

# Data and Preparation

The dataset used in this project comes from NHANES. We restrict the sample to :

-   adults aged over 20
-   from the 2011–2012 wave
-   a subset of variables : `ID, Age, AgeDecade, Gender, Education, MaritalStatus, Race1, nPregnancies, nBabies, Weight, Height, BMI, Pulse, TotChol, Testosterone, PhysActive`

The objective is to clean the data by identifying inconsistencies, applying validation rules, imputing missing values, and ensuring overall coherence.

We begin by filtering the data, removing duplicates, and selecting relevant variables. We also convert factor variables into characters to avoid issues during processing.

```{r}
#| include: false

library(dplyr)
library(NHANES)
library(tidyr)
library(validate)
library(VIM)
library(deductive)
library(errorlocate)
library(ggplot2)

data(NHANES)

df_project <- NHANES %>%
  filter(SurveyYr == "2011_12", Age >= 20) %>%
  distinct(ID, .keep_all = TRUE) %>%
  select(ID, Age, AgeDecade, Gender, Education, MaritalStatus, Race1, nPregnancies, nBabies, Weight, Height, BMI, Pulse, TotChol, Testosterone, PhysActive) %>%
  mutate(across(where(is.factor), as.character))
```

# Validation Rules {.allowframebreaks}

We define a set of validation rules using the `validate` package. These rules allow us to check logical consistency between variables, plausible ranges, and structural relationships. We then evaluate how many observations violate these rules.

We determine some threshold for two variables : 

\footnotesize
```{r}
#| echo: true
#| tidy: true
#| tidy.opts: list(width.cutoff = 40)

# Cholesterol Threshold
res_chol <- summary(df_project$TotChol)
up_chol <- res_chol[["3rd Qu."]] + 1.5 * IQR(df_project$TotChol, na.rm=TRUE)
lo_chol <- res_chol[["1st Qu."]] - 1.5 * IQR(df_project$TotChol, na.rm=TRUE)

# Pulse Threshold
res_pulse <- summary(df_project$Pulse)
up_pulse <- res_pulse[["3rd Qu."]] + 1.5 * IQR(df_project$Pulse, na.rm=TRUE)
lo_pulse <- res_pulse[["1st Qu."]] - 1.5 * IQR(df_project$Pulse, na.rm=TRUE)
```

\normalsize
\framebreak

We then set our rules : 

\footnotesize
```{r}
#| echo: true
#| tidy: true
#| tidy.opts: list(width.cutoff = 40)

rules <- validator(
  age = Age >= 20 & Age <= 80,
  age_decade = is.na(Age) | !is.na(AgeDecade),
  marital_status = !is.na(MaritalStatus),
  education = !is.na(Education),
  physical_activity = !is.na(PhysActive),
  male_not_pregnant = Gender != "male" | (is.na(nPregnancies) & is.na(nBabies)),
  female_pregnant = Gender != "female" | (nPregnancies >= nBabies),
  testosterone = Gender != "male" | Testosterone > 150,
  bmi = abs(BMI - (Weight / (Height/100)^2)) < 0.1,
  pulse = Pulse > lo_pulse & Pulse < up_pulse,
  cholesterol = TotChol > lo_chol & TotChol < up_chol
)

```

# Checking the dataset


```{r}
#| echo: false

rules <- validatetools::remove_redundancy(rules)

cf <- confront(df_project, rules)
check <- summary(cf)
print(check[1:7])

```

# First corrections : rules 2, 3, 4 {.allowframebreaks}

**These rules are about missing values in `AgeDecade` and 2 errors in `MaritalStatus` and `Education`.**

$\rightarrow$ Individuals aged 80 have a missing age decade, which can be reconstructed. 

$\rightarrow$ We use hot-deck imputation to fill missing values in categorical variables. The imputation relies on similar individuals defined by demographic characteristics.

```{r}
#| echo: true

df_project <- df_project %>% 
  mutate(
    AgeDecade_imp = ifelse(Age == 80 & is.na(AgeDecade), TRUE, FALSE),
    AgeDecade = ifelse(AgeDecade_imp == TRUE, "70+", AgeDecade) 
  )

df_project <- hotdeck(
  df_project,
  variable = c("MaritalStatus", "Education"),
  domain_var = c("AgeDecade", "Gender", "Race1"),
  impNA = TRUE 
)

```



```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "80%"
#| out-height: "70%"

aggr(df_project[, c("MaritalStatus", "Education", "MaritalStatus_imp", "Education_imp", "AgeDecade", "AgeDecade_imp")],
     col = c('skyblue', 'tomato', 'green'), 
     delimiter = "_imp", 
     numbers = TRUE, prop = FALSE,
     cex.axis = 0.8, main = "2. Après Hot-deck Démographie")

```

# Bigger corrections : rule 9

**Rule 9 concerns `BMI` variable, which has 24 missing variables.** They are all due to missing variables in `Weight` and `Height`.

$\rightarrow$ We first impute `Weight` and `Height` using hot-deck, based on age and gender. 

$\rightarrow$ Then we compute `BMI` deterministically when it is missing.

```{r}
#| echo: true

df_project <- hotdeck(
  df_project,
  variable = c("Weight", "Height"),
  domain_var = c("AgeDecade", "Gender"),
  impNA = TRUE
)

df_project <- df_project %>%
  mutate(
    BMI_imp = if_else(is.na(BMI), TRUE, FALSE),
    BMI = if_else(BMI_imp == TRUE, Weight / (Height/100)^2, BMI)
  )

```

# Health Variables : rules 8, 10, 11 {.allowframebreaks}

We have identified implausible values such as abnormal pulse or unrealistic cholesterol levels, thanks to our thresholds. **These rules have both errors and missing values.**

```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "80%"
#| out-height: "70%"

df_project %>%
  ggplot(aes(x = Age, y = Pulse, color = Gender)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = lo_pulse, linetype = "dashed", color = "red") + 
  geom_hline(yintercept = up_pulse, linetype = "dashed", color = "red") + 
  labs(title = "Pulse by gender",
       x = "Age",
       y = "Pulse") + 
  theme_minimal() +
  theme(legend.position = "bottom")

```



We directly correct them and impute the errors and missing variables with `knn`.

```{r}
#| echo: true
#| tidy: true
#| tidy.opts: list(width.cutoff = 40)

df_project <- df_project %>% 
  mutate(Testosterone = ifelse(Gender == "male" & Testosterone <= 150, 
                               NA_real_, Testosterone), 
         Pulse = ifelse(!is.na(Pulse) & 
                          (Pulse <= lo_pulse | Pulse >= up_pulse), 
                        NA_real_, Pulse),
         TotChol = ifelse(!is.na(TotChol) & 
                            (TotChol <= lo_chol | TotChol >= up_chol), 
                          NA_real_, TotChol))

df_project <- kNN(
  df_project, 
  variable = c("Pulse", "TotChol", "Testosterone"),
  dist_var = c("Age", "Gender", "Weight", "Height"), 
  k = 5,
  imp_var = TRUE 
)
```

\framebreak
We then check our corrections.

```{r}

cf <- confront(df_project, rules)
check <- summary(cf)
print(check[1:7])

```

Only one rules left to fix. 

# Reproduction Variable : rule 7 {.allowframebreaks}

Reproduction variables require specific constraints. Men cannot have pregnancies, and the number of babies cannot exceed the number of pregnancies. We first enforce these constraints, then re-impute missing or inconsistent values.

```{r}
#| echo: true

df_project <- kNN(
  df_project, 
  variable = c("nPregnancies", "nBabies"),
  dist_var = c("Gender", "Age", "Race1", "Education", "BMI"), 
  k = 5,
  imp_var = TRUE
)
```

\framebreak
But it also imputed pregnancies for men and some women have a number of babies superior to the number of pregnancies. 
```{r}
#| echo: false

cf <- confront(df_project, rules)
check <- summary(cf)
print(check[1:7])
```

We hence force the men to have `NA` and make the imputation find similar duo for `nPregnancies` and `nBabies`. 

```{r}
#| echo: false

df_project <- df_project %>% 
  mutate(
    nPregnancies = ifelse(Gender == "male", NA_integer_, nPregnancies),
    nBabies      = ifelse(Gender == "male", NA_integer_, nBabies),
    nPregnancies = ifelse(Gender == "female" & nPregnancies < nBabies, NA_integer_, nPregnancies),
    nBabies      = ifelse(Gender == "female" & is.na(nPregnancies), NA_integer_, nBabies)
  )

df_project <- hotdeck(
  df_project,
  variable = c("nBabies", "nPregnancies"),
  domain_var = c("AgeDecade", "Gender", "Race1"),
  impNA = TRUE 
)
```

# Final Checks

At the end of the process, we verify that all validation rules are satisfied and that no missing values remain.

```{r}
cf <- confront(df_project, rules)
check <- summary(cf)
print(check[1:7])
```

\centering
**The dataset is now fully cleaned and internally consistent.**

