---
title: "Data Cleaning and Integration in Official Statistics"
subtitle: "Part 2 Project - Donatella Firmani"
author: "Claire du Campe de Rosamel"
format:
  beamer:
    aspectratio: 169
    theme: Madrid
    colortheme: default
    fonttheme: professionalfonts
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(palmerpenguins)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(knitr)

set.seed(123)

penguins_clean <- penguins %>%
  select(
    species,
    island,
    sex,
    flipper_length_mm,
    body_mass_g,
    bill_length_mm
  ) %>%
  na.omit()

n <- nrow(penguins_clean)

sample_A_id <- sample(
  1:n,
  size = floor(n / 2),
  replace = FALSE
)

A_full <- penguins_clean[sample_A_id, ]
B_full <- penguins_clean[-sample_A_id, ]

A <- A_full %>%
  select(
    species,
    island,
    sex,
    flipper_length_mm,
    body_mass_g
  )

B <- B_full %>%
  select(
    species,
    island,
    sex,
    flipper_length_mm,
    bill_length_mm
  )

A_true_bill <- A_full$bill_length_mm
```

```{r functions}
hot_deck_match <- function(recipient_file, donor_file, spec = "rich") {
  
  imputed_values <- numeric(nrow(recipient_file))
  
  for (i in 1:nrow(recipient_file)) {
    
    recipient <- recipient_file[i, ]
    
    if (spec == "rich") {
      possible_donors <- donor_file %>%
        filter(
          species == recipient$species,
          sex == recipient$sex,
          island == recipient$island
        )
    }
    
    if (spec == "poor") {
      possible_donors <- donor_file %>%
        filter(
          species == recipient$species
        )
    }
    
    if (nrow(possible_donors) == 0) {
      possible_donors <- donor_file %>%
        filter(species == recipient$species)
    }
    
    if (nrow(possible_donors) == 0) {
      possible_donors <- donor_file
    }
    
    if (spec == "rich") {
      best_donor <- possible_donors %>%
        mutate(
          distance = abs(flipper_length_mm - recipient$flipper_length_mm)
        ) %>%
        slice_min(distance, n = 1, with_ties = TRUE) %>%
        slice_sample(n = 1)
    }
    
    if (spec == "poor") {
      best_donor <- possible_donors %>%
        slice_sample(n = 1)
    }
    
    imputed_values[i] <- best_donor$bill_length_mm
  }
  
  recipient_file %>%
    mutate(bill_length_imputed = imputed_values)
}

evaluate_matching <- function(A_file, B_file, A_true_values, spec = "rich") {
  
  matched_data <- hot_deck_match(A_file, B_file, spec = spec) %>%
    mutate(
      bill_length_true = A_true_values,
      error = bill_length_imputed - bill_length_true
    )
  
  true_cor <- cor(
    matched_data$body_mass_g,
    matched_data$bill_length_true,
    use = "complete.obs"
  )
  
  synthetic_cor <- cor(
    matched_data$body_mass_g,
    matched_data$bill_length_imputed,
    use = "complete.obs"
  )
  
  tibble(
    specification = spec,
    rmse = sqrt(mean(matched_data$error^2, na.rm = TRUE)),
    mae = mean(abs(matched_data$error), na.rm = TRUE),
    true_correlation = true_cor,
    synthetic_correlation = synthetic_cor,
    correlation_error = abs(true_cor - synthetic_cor)
  )
}

A_matched <- hot_deck_match(A, B, spec = "rich") %>%
  mutate(
    bill_length_true = A_true_bill,
    error = bill_length_imputed - bill_length_true
  )

baseline_results <- evaluate_matching(
  A_file = A,
  B_file = B,
  A_true_values = A_true_bill,
  spec = "rich"
)
```

## Research question

This project applies **statistical matching** to the Palmer Penguins dataset.

The objective is to reconstruct missing joint information between:

- **body mass**, observed only in file A;
- **bill length**, observed only in file B.

\vspace{0.3cm}

Main question:

$$
\text{Can hot deck matching recover the relationship between body mass and bill length?}
$$

## Building a statistical matching problem

The original dataset is complete.

To reproduce a statistical matching setting, it is split in two ways:

- **horizontally:** penguins are divided into two disjoint files;
- **vertically:** some variables are kept only in one file.

```{r}
tibble(
  File = c("A", "B"),
  Common_variables = c(
    "species, island, sex, flipper length",
    "species, island, sex, flipper length"
  ),
  Observed_specific_variable = c("body mass", "bill length"),
  Hidden_variable = c("bill length", "body mass")
) %>%
  kable()
```

## Statistical matching design

In file A, bill length is treated as missing.

In file B, bill length is observed and can be used for donation.

\vspace{0.2cm}

For each penguin in A:

1. search for similar penguins in B;
2. select one donor;
3. assign the donor's bill length to the recipient penguin.

\vspace{0.2cm}

The output is a synthetic version of file A:

$$
X + Y + \widehat{Z}
$$

where X are common variables, Y is body mass, and $\widehat{Z}$ is imputed bill length.

## Baseline method: rich distance hot deck

The main procedure is the most informative specification.

For each recipient in A, donors are first restricted to the same:

- species;
- sex;
- island.

Among these possible donors, the selected donor is the one with the closest flipper length.

$$
d(i,j) = |flipper\_length_i - flipper\_length_j|
$$

This is the **baseline statistical matching procedure**.

## Code excerpt: main matching rule

```{r, echo=TRUE, eval=FALSE, fig.align="center"}
best_donor <- donor_file %>%
  filter(
    species == recipient$species,
    sex == recipient$sex,
    island == recipient$island
  ) %>%
  mutate(
    distance = abs(flipper_length_mm - recipient$flipper_length_mm)
  ) %>%
  slice_min(distance, n = 1, with_ties = TRUE) %>%
  slice_sample(n = 1)
```

\vspace{0.2cm}

This code is the core of the rich distance hot deck procedure.

## Baseline evaluation

Because the experiment starts from a complete dataset, the hidden values are known.

This allows a direct evaluation of the imputation.

```{r}
baseline_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable()
```

The baseline is evaluated using individual errors and preservation of the relationship between Y and Z.

## Distribution of true and imputed bill length

```{r, out.height="70%", out.width="80%", fig.align="center"}
A_matched %>%
  select(bill_length_true, bill_length_imputed) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "bill_length_mm"
  ) %>%
  ggplot(aes(x = bill_length_mm, linetype = variable)) +
  geom_density(linewidth = 1) +
  labs(
    title = "True vs imputed distribution of bill length",
    x = "Bill length (mm)",
    y = "Density",
    linetype = ""
  ) +
  theme_minimal()
```

## Reconstructed relationship between Y and Z

```{r, out.height="70%", out.width="80%", fig.align="center"}
A_matched %>%
  select(body_mass_g, bill_length_true, bill_length_imputed) %>%
  pivot_longer(
    cols = c(bill_length_true, bill_length_imputed),
    names_to = "variable",
    values_to = "bill_length_mm"
  ) %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable) +
  labs(
    title = "True vs synthetic relationship",
    x = "Body mass (g)",
    y = "Bill length (mm)"
  ) +
  theme_minimal()
```

## Sensitivity analysis

After applying the baseline procedure, we test whether matching quality depends on:

- the variables used to define similarity;
- the available sample size.

\vspace{0.2cm}

Two specifications are compared:

```{r}
tibble(
  Specification = c("Poor", "Rich"),
  Rule = c(
    "Random donor with the same species",
    "Same species, sex and island, then nearest flipper length"
  )
) %>%
  kable()
```

## Fast simulation design

```{r simulation}
run_one_simulation <- function(sample_size, specification) {
  
  sampled_data <- penguins_clean %>%
    slice_sample(n = sample_size)
  
  sample_A_id <- sample(
    1:nrow(sampled_data),
    size = floor(sample_size / 2),
    replace = FALSE
  )
  
  A_full_sim <- sampled_data[sample_A_id, ]
  B_full_sim <- sampled_data[-sample_A_id, ]
  
  A_sim <- A_full_sim %>%
    select(
      species,
      island,
      sex,
      flipper_length_mm,
      body_mass_g
    )
  
  B_sim <- B_full_sim %>%
    select(
      species,
      island,
      sex,
      flipper_length_mm,
      bill_length_mm
    )
  
  evaluate_matching(
    A_file = A_sim,
    B_file = B_sim,
    A_true_values = A_full_sim$bill_length_mm,
    spec = specification
  ) %>%
    select(rmse, mae, correlation_error)
}

simulation_results <- expand_grid(
  sample_size = c(50, 100, 150, 250, 300),
  repetition = 1:20,
  specification = c("poor", "rich")
) %>%
  mutate(
    result = pmap(
      list(sample_size, specification),
      ~ run_one_simulation(
        sample_size = ..1,
        specification = ..2
      )
    )
  ) %>%
  unnest(result)

simulation_summary <- simulation_results %>%
  group_by(sample_size, specification) %>%
  summarise(
    mean_rmse = mean(rmse, na.rm = TRUE),
    mean_mae = mean(mae, na.rm = TRUE),
    mean_correlation_error = mean(correlation_error, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, echo=TRUE, eval=FALSE}
simulation_results <- expand_grid(
  sample_size = c(50, 100, 150, 250, 300),
  repetition = 1:20,
  specification = c("poor", "rich")
)
```

The simulation is deliberately limited to keep computation time reasonable.

## Matching quality and sample size

```{r, out.height="70%", out.width="80%", fig.align="center"}
ggplot(
  simulation_summary,
  aes(x = sample_size, y = mean_rmse, linetype = specification)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Matching quality by sample size",
    x = "Total sample size before splitting into A and B",
    y = "Average RMSE",
    linetype = "Specification"
  ) +
  theme_minimal()
```


## Main findings

The baseline rich distance hot deck uses the most informative available common variables.

The evaluation compares the synthetic information to the true hidden values.

The sensitivity analysis shows whether results depend on matching assumptions and donor file size.

\vspace{0.2cm}

Overall, hot deck matching can reconstruct part of the missing joint information, but its quality depends on the informativeness of X.

## Conclusion

This project illustrates a micro non-parametric statistical matching approach.

The Palmer Penguins dataset is useful because the true hidden values are known before the artificial split.

The main limitation is that, in a real application, the true missing values would not be observed.

\vspace{0.2cm}

The method is therefore useful, but it relies on a strong practical condition:

$$
X \text{ must be informative enough about both } Y \text{ and } Z.
$$

